---
title: "Output tables with `mlth.data.frame`"
author: "Ivan Voronin"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
vignette: >
  %\VignetteIndexEntry{Output tables with `mlth.data.frame`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Hi!

This vignette illustrates the functionality of `mlth.data.frame` package. This name stands for __multiheaded data.frame__, a data.frame with grouped columns.

This package is a companion of [`TwinAnalysis` package](https://github.com/IvanVoronin/TwinAnalysis). As such, it provides functionality for organizing and collecting the tables across the analysis script and sending them into an Excel file or on a html/markdown/LaTeX report. 

To install the package:

```{r, eval = FALSE}
# install.packages('devtools')
devtools::install_github('IvanVoronin/mlth.data.frame')

library(mlth.data.frame)
```

```{r, echo = FALSE}
# Example here
library('mlth.data.frame')
```

A mlth.data.frame looks like this:

```{r}
A <- mlth.data.frame(Letters = list(cons = c('a', 'e', 'i'),
                                    vow = c('b', 'c', 'd')),
                     Numbers = list(even = c(2, 4, 6),
                                    odd = c(1, 3, 5)))

A
```

This mlth.data.frame has `r ncol(A)` columns and `r nrow(A)` rows. Use `ncol()`, `nrow()` and `dim()` to get the size of the table:

```{r}
ncol(A)

nrow(A)

dim(A)
```

The columns of mlth.data.frame can be arranged into groups, in this case Letters and Numbers. You can select groups and variables by refering to them by name or index number:

```{r}
A$Letters

A[['Numbers']]

# Mind the difference between `[` and `[[`
A['Numbers']
```

If you select a single column, it will return vector, otherwise - mlth.data.frame.

```{r}
A$Letters$cons

# But:
A$Letters['cons']

class(A$Letters['cons'])
```

Any mlth.data.frame is a list, so all appropriate functions can be applied to them.

```{r}
names(A)

lapply(A, names)

rapply(A, class)
```


I will use the in-built dataset `mtcars` to demonstrate the functionality of mlth.data.frame for the output. I want to compare the cars with 4, 6 and 8 cylinders (`cyl`) by mileage (`mpg`), horsepower (`hp`) and weight. This is one of many possible ways to do so and get the result in mlth.data.frame.

```{r}
data(mtcars)
library(dplyr)

mtcars$cyl <- factor(mtcars$cyl,
                     levels = c(4, 6, 8),
                     labels = paste(c(4, 6, 8), 'cylinders'))

var_names <- c('mpg', 'hp', 'wt')

Tab <- mtcars[, var_names] %>%
  by(mtcars$cyl, function(x) {
    lapply(x, function(y) {
      y <- na.omit(y)
      data.frame(n = length(y), 
           M = mean(y),
           SD = sd(y))
    }) %>% do.call('rbind', .)
  }) %>% do.call('mlth.data.frame', .) %>%
  { row.names(.) <- var_names; . }

Tab
```

Now let's append the results from ANOVA.

```{r}
Tab <- mtcars[, var_names] %>%
  lapply(function(x) {
    LM <- summary(lm(x ~ mtcars$cyl))
    ftest <- LM$fstatistic
    data.frame(F = ftest['value'],
               df1 = ftest['numdf'],
               df2 = ftest['dendf'],
               p = pf(ftest['value'], 
                      ftest['numdf'], 
                      ftest['dendf'], 
                      lower.tail = FALSE),
               R2 = LM$r.squared)
  }) %>% do.call('rbind', .) %>%
  as.mlth.data.frame %>%
  cbind(Tab, .)

Tab
```

To send this table to an Excel output, we have to register it. Then, at the end of the script, we will use `write.xlsx.output()` to write all the tables to a single Excel file. The table will be appended to the global `OUTPUT` list. We can point out the name of Excel spreadsheet in the file (`name`), table caption (`caption`) and table footnote (`note`). Details in `?register_output`. The table doesn't have to be mlth.data.frame, the regular data.frame can also be recorded for the output. 

```{r}
register_output(Tab, name = 'by cylinders',
                caption = 'Characteristics of the cars with different engines')
```


```{r}
# This is an example of adding a data.frame to the output
data.frame(A = 1:3, B = 4:6) %>%
  register_output(name = 'lonely data.frame')
# It works
```

To forward the table to Rmarkdown document, use `knitr` and `kableExtra` ( [more about kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)). The latter package provides additional functionality for table formating, including layered headers. Here is an example of formating the table as html.

```{r}
library(knitr)
library(kableExtra)

Tab %>% behead %>%
  kable(digits = c(0, 2, 2, 0, 2, 2, 0, 2, 2, 3, 0, 0, 3, 3))  %>% 
  add_complex_header_above(Tab) %>%
  kable_styling
```

At the end we write all the output into a file. Currently, `openxlsx` package is used to write the tables into xlsx file. But you can define your own writer function based on any other package of your choice and pass it to `write.xlsx.output()`. Details in `?write.xlsx.output` and `?xlsx.writer.openxlsx`.

```{r}
write.xlsx.output('vignette_output.xlsx')
```