---
title: "Output tables with `mlth.data.frame`"
author: "Ivan Voronin"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
vignette: >
  %\VignetteIndexEntry{Output tables with `mlth.data.frame`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Hi!

This vignette illustrates the functionality of `mlth.data.frame` package. This name stands for __multiheaded data.frame__, a data.frame with grouped columns. It also allows grouped rows when
printing using `kable` or writing in an Excel spreadsheet.

This package is a companion of [`TwinAnalysis` package](https://github.com/IvanVoronin/TwinAnalysis). As such, it provides functionality for organizing and collecting the tables across the analysis script and sending them into an Excel file or on a html/markdown/LaTeX report. 

# Installation

```{r, eval = FALSE}
# install.packages('devtools')
devtools::install_github('IvanVoronin/mlth.data.frame')

library(mlth.data.frame)
```

```{r, echo = FALSE}
# Example here
library('mlth.data.frame')
```

# Basics

```{r}
A <- mlth.data.frame(
  Letters = list(
    cons = c('a', 'e', 'i'),
    vow = c('b', 'c', 'd')),
  Numbers = list(
    even = c(2, 4, 6),
    odd = c(1, 3, 5)
  )
)

A
```

This mlth.data.frame has `r ncol(A)` columns and `r nrow(A)` rows. Use `ncol()`, `nrow()` and `dim()` to get the size of the table:

```{r}
ncol(A)

nrow(A)

dim(A)
```

The columns of mlth.data.frame can be arranged into groups, in this case Letters and Numbers. You can select groups and variables by referring to them by name or index number:

```{r}
A$Letters

A[['Numbers']]

# Mind the difference between `[` and `[[`
A['Numbers']
```

If you select a single column, it will return vector, otherwise - mlth.data.frame.

```{r}
A$Letters$cons

# But:
A$Letters['cons']

class(A$Letters['cons'])
```

Any mlth.data.frame is a list, so all appropriate functions can be applied to them.

```{r}
names(A)

lapply(A, names)

rapply(A, class)
```

# `mlth.data.frame` for output

I use the in-built data `mtcars` to show the functionality of mlth.data.frame for the output. I compare the cars with 4, 6 and 8 cylinders (`cyl`) by mileage (`mpg`), horsepower (`hp`) and weight. This is one of many possible ways to do so and get the result in mlth.data.frame.

```{r}
data(mtcars)
library(dplyr)

mtcars$cyl <- factor(
  mtcars$cyl,
  levels = c(4, 6, 8),
  labels = paste(c(4, 6, 8), 'cylinders')
)

var_names <- c('mpg', 'hp', 'wt')

Tab <- mtcars[, var_names] %>%
  by(mtcars$cyl, function(x) {
    lapply(x, function(y) {
      y <- na.omit(y)
      data.frame(n = length(y), 
           M = mean(y),
           SD = sd(y))
    }) %>% do.call('rbind', .)
  }) %>% do.call('mlth.data.frame', .) %>%
  { row.names(.) <- var_names; . }

Tab
```

Now let's append the results from ANOVA.

```{r}
Tab <- mtcars[, var_names] %>%
  lapply(function(x) {
    LM <- summary(lm(x ~ mtcars$cyl))
    ftest <- LM$fstatistic
    data.frame(
      F = ftest['value'],
      df1 = ftest['numdf'],
      df2 = ftest['dendf'],
      p = pf(
        ftest['value'], 
        ftest['numdf'], 
        ftest['dendf'], 
        lower.tail = FALSE
      ),
      R2 = LM$r.squared
    )
  }) %>% do.call('rbind', .) %>%
  as.mlth.data.frame %>%
  cbind(Tab, .)

Tab
```

# Writing html tables using `knitr::kable` and `kableExtra`

```{r}
# library(knitr)
library(kableExtra)

Tab %>%
  behead %>%
  kable(
    align = 'c',
    digits = c(
      rep(c(0, 2, 2), 3),
      3, 0, 0, 3, 3
    )
  ) %>%
  kable_styling %>%
  add_complex_header_above(Tab)
```


```{r include=FALSE}
Tab[1:3] %>%
  lapply(as.data.frame) %>%
  lapply(`row.names<-`, c('mpg', 'hp', 'wt')) %>%
  kable2(
    align = 'c',
    align_first = 'l',
    digits = c(0, 2, 2),
    register_output = TRUE,
    caption = 'A table',
    footnote = 'Footnote for this table'
  ) %>%
  kable_styling

list(
  'Table 1' = Tab,
  'Table 2' = Tab
) %>%
  kable2 %>%
  kable_styling()
```


# Writing the tables to an Excel spreadsheet

To send this table to an Excel output, we have to register it. Then, at the end of the script, we will use `write.xlsx.output()` to write all the tables to a single Excel file. The table will be appended to the global `OUTPUT` list. We can point out the name of Excel spreadsheet in the file (`name`), table caption (`caption`) and table footnote (`note`). Details in `?register_output`. The table doesn't have to be mlth.data.frame, the regular data.frame can also be recorded for the output. 

```{r}
register_output(
  Tab, #name = 'by cylinders',
  caption = 'Characteristics of the cars with different engines'
)
```


```{r}
# This is an example of adding a data.frame to the output
data.frame(A = 1:3, B = 4:6) %>%
  register_output(
    #name = 'lonely data.frame'
  )
# It works
```

To forward the table to Rmarkdown document, use `knitr` and `kableExtra` ( [more about kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)). The latter package provides additional functionality for table formating, including layered headers. Here is an example of formating the table as html.

```{r include=FALSE}
# library(knitr)
library(kableExtra)

A <- mlth.data.frame(
  A = list(
    AA = 1:3,
    AB = list(
      ABA = 1:3,
      ABB = 1:3
    )
  ),
  B = list(
    BA = list(
      BAA = 1:3,
      BAB = 1:3
    ),
    BB = list(
      BBA = 1:3,
      BBB = 1:3
    )
  )
)

A %>% behead

A %>% kable2 %>% kable_styling

Tab %>% behead %>%
  kable(digits = c(0, 2, 2, 0, 2, 2, 0, 2, 2, 3, 0, 0, 3, 3))  %>% 
  add_complex_header_above(Tab) %>%
  kable_styling
```


```{r}
data.frame(
  A = 1:3,
  B = 4:6
) %>%
  kable2 %>%
  kable_styling

row.names(Tab) <- 1:3

Tab[['4 cylinders']] <- 1:3

Tab %>%
  kable2(
    row.names = TRUE
  ) %>%
  kable_styling


A <- mlth.data.frame(
  n = 1:3,
  m = 3:5,
  L = list(
    k = 4:6,
    r = 2:4
  )
)

A %>%
  kable2(
    align_first = c('l', 'l'),
    align = 'c'
  ) %>%
  kable_styling
```


```{r}
data.frame(
  A = 1:3,
  B = letters[1:3]
) %>%
  register_output(
    caption = 'Another data.frame'
  )

mlth.data.frame(
  A = 1:3,
  L = list(
    B = letters[1:3],
    C = LETTERS[1:3]
  )
)%>%
  register_output(
    caption = 'More data.frames'
  )

```

At the end we write all the output into a file. Currently, `openxlsx` package is used to write the tables into xlsx file. But you can define your own writer function based on any other package of your choice and pass it to `write.xlsx.output()`. Details in `?write.xlsx.output` and `?xlsx.writer.openxlsx`.

```{r}
write.xlsx.output('vignette_output.xlsx')
```

```{r}
# Testing row.names
A <- data.frame(
  X1 = 1:3,
  X2 = 4:6
#  row.names = letters[1:3]
)
B <- data.frame(
  Y1 = 7:9,
  Y2 = 10:12
#  row.names = letters[1:3]
)
mlth.data.frame(A = A, B = B)
```